Compression-based inference on graph data

### 2013-03-17 Benelearn/src/benelearn.tex
...

\usepackage{style/benelearn2013}
\usepackage{charter}
\usepackage{eulervm}
% \usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\usepackage{subcaption}

@@

@@

\begin{abstract}
We investigate the use of compression-based learning on graph data. We hypothesize that the sequential representation of data on which general purpose compressors operate, hampers compression-based learning, as a graph can be represented sequentially in many different, equivalent ways. Using Normalized Compression Distance (NCD), we test various compressors that operate directly on the graph representation against compressors that operate on a bitstring representation. We use both synthetic, randomly generated graphs and real-life datasets. 
\end{abstract}

@@

The idea of compression-based learning is expressed in two strongly related fields: \emph{Minimum Description Length} and \emph{Kolmogorov Complexity}. Minimum Description Length \cite{}@! builds a statistical framework on the principle that a good model is one that can be used to describe the data in as few bits as possible. Kolmogorov Complexity is concerned with the mathematical expression of the \emph{information content} of a piece of data. If we can find some short description of a dataset (ie. compress it) then the total information content must be below the length of that description.

@@

From the many methods born from Kolmogorov Complexity and Minimum Description Length, we choose \emph{Normalized Compression Distance} (NCD) as a representative method of compression-based learning. The choice is a practical one, NCD is a simple method which requires nothing more than a compressor fit for the type of data under question. Any domain-specific knowledge we wish to use about our data (eg. it represents a graph) must be placed in the compressor. We proceed under the assumption that conclusions reached about the performance of NCD on graph data will translate to MDL and other frameworks.

We will provide a brief, intuitive explanation of the principles involved, sufficient to understand the ideas presented in this paper. For a more in-depth and rigorous treatment we refer the reader to @! and @!.

@@

Kolmogorov complexity is a notion of information content, based on two principles: (a) all data can be represented as a bitstring (b) the shorter this bitstring can be described, the less information is contained in it.

@@

This intuition prompted Li and Vitanyi @! \cite{} to investigate the use of Kolmogorov Complexity as a metric. The first issue is that $K(\cdot\mid \cdot)$ is not symmetric. It takes a small program to blank out the collected works of Shakespeare, but the reverse is a more complex operation. The first step is thus to define the (symmetric) Information Distance:
\[
ID(x, y) = \max \left [K(x\mid y),K(y\mid x) \right ] 
\]

@@

@@

Our aim is to test a selection of compressors on a variety of graph data. To ascertain the performance of the compressors, we generate graphs from different sources, calculate their distances and see whether a clustering algorithm can reconstruct the original sources as clusters.

@@

We generate graphs from two models. The first is the classic Erd\H{o}s-R\'enyi model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert model \cite{}@!, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to form a scale free distribution.

Since we want there to be some challenge in separating the two classes of network, we, ensure that they have the same number of nodes and links. To achieve this, we first generate the BA networks, count their nodes and links and use these as parameters for the ER model.

Once we have this dataset, calculate the NCD with a given compressor for every pair of graphs in the dataset, giving us a symmetric matrix. We use the k-medioid algorithm to cluster this set into two clusters, and we check whether graphs from a given source tend to end up in a specific cluster.

@@

We use the following datasets:
\begin{description}
	\item[web]
	\item[metabolic] 
	\item[semantic A] A semantic description of 
	\item[semantic B]
\end{description}

Some basic properties of these graphs are given in tabel \ref{table:properties}.

@@

Subdue \cite{} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches through all possible subgraphs, searching for the  one that maximally compresses the data.

@@

For larger graphs, the problem becomes easier (as there is more structure), but we can add the fractal generator. This generator can only produce larger graphs, and with its inclusion we can no longer ensure that all graphs have equal size. 
\begin{table}[h]
\label{table:synthetic-large}
\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world) & @!/@!  & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{GZIP: error @!}
\end{subfigure}
\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world)& @!/@! & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{Subdue: error @!}
\end{subfigure}
\caption{\ldots}
\end{table}

@@

\begin{table}[h]
\label{table:real-life}
\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world) & @!/@!  & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{GZIP: error @!}
\end{subfigure}
\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world)& @!/@! & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{Subdue: error @!}
\end{subfigure}
\caption{\ldots}
\end{table}

@@

This result suggests that inference on graphs can be performed by sequential algorithms on a sequential representation in linear time, with decent results.

@@

>>>>

\usepackage{style/benelearn2013}
\usepackage{charter}
\usepackage{eulervm}
% \usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}

@@

\newfloat{pseudo}{h}{lop}
\floatname{pseudo}{Algorithm}

@@

\begin{abstract}
We investigate the use of compression-based learning on graph data. Our hypothesis is that the sequential representation of data on which general purpose compressors operate, hampers compression-based learning, as a graph can be represented sequentially in many different, equivalent ways. Using Normalized Compression Distance (NCD), we test various compressors that operate directly on the graph representation against compressors that operate on a bitstring representation. We use both synthetic, randomly generated graphs and real-life datasets. 
\end{abstract}

@@

The idea of compression-based learning is expressed in two related fields: \emph{Minimum Description Length} and \emph{Kolmogorov Complexity}. Minimum Description Length \cite{grunwald2007minimum} builds a statistical framework on the principle that a good model is one that can be used to describe the data in as few bits as possible. Kolmogorov Complexity \cite{li1997introduction} is concerned with the mathematical expression of the \emph{information content} of  data. If we can find some short description of a dataset (ie. compress it) then the total information content must be below the length of that description.

@@

From the many methods born from Kolmogorov Complexity and Minimum Description Length, we choose \emph{Normalized Compression Distance} (NCD) as a representative method of compression-based learning. The choice is a practical one: NCD is a simple method which requires nothing more than a general purpose compressor. Any domain-specific knowledge we wish to use about our data (eg. it represents a graph) can be placed in the compressor. We proceed under the assumption that conclusions reached about the performance of NCD on graph data will translate to MDL and other frameworks.

We will provide a brief, intuitive explanation of the principles involved, sufficient to understand the ideas presented in this paper. For a more in-depth and rigorous treatment we refer the reader to \cite{li2004similarity} and \cite{cilibrasi2005clustering}. For a general introduction to Kolmogorov complexity, see \cite{li1997introduction}.

@@

Kolmogorov complexity is a notion of information content, based on two principles: (a) all data can be represented as a bitstring (b) the shorter this string can be described, the less information is contained in it.

@@

This intuition prompted Li and Vit{\'a}nyi \cite{li2004similarity} to investigate the use of Kolmogorov Complexity as a metric. The first issue is that $K(\cdot\mid \cdot)$ is not symmetric: it takes a small program to blank out the collected works of Shakespeare, but the reverse is a more complex operation. The first step is thus to define the (symmetric) Information Distance:
\[
ID(x, y) = \max \left [K(x\mid y),K(y\mid x) \right ] 
\]

@@

When our data is represented as a graph, rather than a string, we replace the notion of concatenation strings by concatenation of graphs. That is, we combine the graphs $x$ and $y$ into a single (disconnected) graph.

@@

Our aim is to test a selection of compressors on a variety of graph data. To ascertain the performance of the compressors, we generate graphs from different sources, calculate their distances and see whether a clustering algorithm can reconstruct the original sources as clusters. Source code for these experiments is available \footnote{@!@!}

@@

We generate graphs from two models. The first is the classic Erd\H{o}s-R\'enyi (ER) model, where a uniform random choice is made from all graphs with $n$ nodes and $m$ links. The second is the Barab\'asi-Albert (BA) model \cite{albert2002statistical}, which grows a graph from a set of $n_0$ unconnected nodes, one node at a time, connecting each new node to $k$ distinct existing nodes where the probability that a given existing node is chosen for a connection is its degree, divided by the sum of the degrees of all nodes. Thus, under the BA model the more links a node has, the higher the probability that it will accrue even more. This effect causes the degree distribution of a BA network to form a scale free distribution.

Since we want there to be some challenge in separating the two classes of network, we ensure that they have the same number of nodes and links. To achieve this, we first generate the BA networks, count their nodes and links and use these as parameters for the ER model.

Once we have this dataset, calculate the NCD with a given compressor for every pair of graphs in the dataset, giving us a symmetric matrix. We use the k-medoids algorithm to cluster this set into two clusters, and we check whether graphs from a given source tend to end up in a specific cluster.

For second experiment with larger graphs, we can include the fractal graph generation algorithm from \cite{song2006origins}. We set the hub-parameter which determines the level of fractality (as a trade-off with the level of small worldness) to $0.1$ (for a small world network) and to $0.9$ (for a mostly fractal network).

\subsubsection*{Parameter settings}

@!

For all graphs we shuffle the node order before adding them to the dataset.

@@

We use the following datasets:
\begin{description}
	\item[web] A graph snapshot of a section of the world wide web. All nodes and edges have the same label. (This is roughly the same as not having labels, in terms of compression. To make the compressor as general as possible we choose some canonical string as the label for all nodes and links).
	\item[metabolic] A metabolic network. @!
	\item[semantic A] A semantic description of researchers at the @! institute \ldots 
	\item[semantic B]
\end{description}

Some basic properties of these graphs are given in table \ref{table:properties}.

@@

Subdue \cite{ketkar2005subdue} is an algorithm for finding frequent subgraphs in graph data. The algorithm searches through all possible subgraphs, searching for the  one that maximally compresses the data. The algorithm is essentially a beam search through the space of subgraphs.

Each subgraph serves as a model for compression by finding all occurrences of the subgraph in the data. Each occurrence is replaced by a single node, with the edges to the nodes containing annotations on how the edge shouldbe connected within the subgraph. We then store the subgrap once, together with the edited datagraph, to represent the original data. If the subgraph occurs often enough to offset the extra bits required to annotate the edges in the datagraph, the Subdue representation will be more efficient.

The following is a rough description of the algorithm:

\begin{pseudo}[h]
{
	$G$: the data graph \\
	$b$: the beam size \\
	\\
	$S \leftarrow [K_1]$ \textit{\#  initialize the list of substructures with a graph of a single node} \\
	\\
	\textbf{loop} \\
	\hspace*{5mm} $s \leftarrow $ \textit{head element of} $S$ \\
	\hspace*{5mm} \textit{add all extensions of} $s$ \textbf{to} $S$ \\

	\hspace*{5mm}\textit{sort $S$ by score($s', G$)} for $s' \in S$\\
	\hspace*{5mm}\textit{remove all but the first $b$ elements of $S$} \\
	\\
	\textbf{function } $(s', G)$\\
	\hspace*{5mm} \textit{replace occurrences of $s'$ in $G$ with node $N$}\\
	\hspace*{5mm} \textit{annotate links to $N$ with the nodes in $s'$}\\
	\hspace*{5mm} \textbf{return} \textit{number of bits to store the edited $G$ and $s'$ }\\
}
	\caption{Pseudocode for the Subdue algorithm}
\end{pseudo}

@@

For larger graphs, the problem becomes easier (as there is more structure), but we can add the fractal generator. This generator can only produce larger graphs, and with its inclusion we can no longer ensure that all graphs have equal size. 
\begin{table}[h]
\label{table:synthetic-large}
\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world) & @!/@!  & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{GZIP: error @!}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\begin{tabular}{l | r | r r r r}
\hline
  ER      & @!/@!  & @! & @! & @! & @! \\
  BA      & @!/@!  & @! & @! & @! & @!\\
  fractal (pure) & @!/@!  & @! & @! & @! & @!\\
  fractal (small world)& @!/@! & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{Subdue: error @!}
\end{subfigure}
\caption{\ldots}
\end{table}

@@

\begin{table}[h]
\label{table:real-life}
\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | r | r r r r}
\hline
  Web      & @!/@!  & @! & @! & @! & @! \\
  Metabolic     & @!/@!  & @! & @! & @! & @!\\
  Semantic A & @!/@!  & @! & @! & @! & @!\\
  Semantic B & @!/@!  & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{GZIP: error @! (@!)}
\end{subfigure}
\vspace{3mm}

\begin{subfigure}[b]{1\columnwidth}
\centering
\begin{tabular}{l | r | r r r r}
\hline
  Web      & @!/@!  & @! & @! & @! & @! \\
  Metabolic     & @!/@!  & @! & @! & @! & @!\\
  Semantic A & @!/@!  & @! & @! & @! & @!\\
  Semantic B & @!/@! & @! & @! & @! & @!\\
\hline
\end{tabular}
\caption{Subdue: error @! (@!)}
\end{subfigure}
\caption{\ldots}
\end{table}

@@

This result suggests that inference on graphs can be performed by sequential algorithms on a sequential representation in linear time, with decent results, depending on the complexity of the task.

@@

\bibliographystyle{plain} @!
\bibliography{benelearn}
